---
output: 
  html_document:
    code_download: true
    includes:
      after_body: footer.html
---

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.showtext = TRUE)

library(tidyverse)
library(tidytuesdayR)
library(patchwork)
library(showtext)

font_add_google(name = "Oswald")

showtext_auto()
theme_set(theme_minimal())
```

## **TidyTuesday: Global Deforestation**

*Posted on April 6, 2021*

A quick analysis of the weekly [\#TidyTuesday](http://github.com/rfordatascience/tidytuesday) dataset organized by the R4DS Online Learning Community. My approach is to apply my data science skills to explore one question I have about the data and generate a visualization that addresses this question. The main purpose for me is to practice and try out new things. I am never completely satisfied with the end result but I do the best I can in a short period of time.

**What I learned this week about R and the Tidyverse**

-   Using a highlight color in a bar chart to focus attention on certain effects
-   Using a custom axis label using the ```label_number``` function from the ```scales``` package

**Brief explanation of the dataset**

For this #TidyTuesday, R4DS provided five datasets with two about forest area change by country, one about specific causes of forest loss in Brazil, one about soybean production and one about vegetable oil production.  I was interested in the forest change by year and country but neither dataset gave me exacttly what I was looking for.  I went back to the source from [OurWorldInData](https://ourworldindata.org/forests-and-deforestation) to get the complete dataset by year and country.  I then drilled down on by continent and by country.  Finally, I analyzed the causes for deforestation in Brazil.

```{r Load, include = FALSE}
#Save needed data into rds for blog post
load("./data/tt_210406.Rdata")
```

### Wrangle

Initial conclusions from exploring the forest_area_km dataset:

* The meaning of Northern America was not clear.  Upon further investigation, the continent of North America included Northern America, Central America and Caribbean entities.  
* Forest area data was available from 1990 through 2020.  
* Causes for deforestation and forest area loss in Brazil were given from 2001 to 2013.  

Summarize the forest area change between 1990 and 2020 for each of the continents.  As mentioned above, three entities need to be recoded for North America.
```{r}
continents_plus <- c("Africa", "Asia", "Europe", "Northern America", "South America", "Oceania", "Central America", "Caribbean")

forest_continent <- forest_area_km %>%
  filter(entity %in% continents_plus) %>%
  mutate(entity = case_when(entity == "Northern America" ~ "North America",
                            entity == "Central America" ~ "North America",
                            entity == "Caribbean" ~ "North America",
                            TRUE ~ entity)) %>%
  group_by(entity, year) %>%
  summarize(forest_area = sum(forest_area), .groups = "drop")

forest_change <- forest_continent %>%
  filter(year == 1990 | year == 2020) %>%
  pivot_wider(names_from = year, names_prefix = "forest_area_", values_from = forest_area ) %>%
  mutate(forest_change = forest_area_2020 - forest_area_1990)
```

### Visualize

First, let's look at which continents are seeing the largest changes in forest area.  A custom color palette is used to highlight the biggest contributor to deforestation: South America.
```{r}
single_highlight_pal <- c("Asia" = "#808080", "Europe" = "#808080", "North America" = "#808080", 
                        "Oceania" = "#808080", "Africa" = "#808080", "South America" = "#FFFF00")

p1 <- forest_change %>%
  mutate(entity = fct_reorder(entity, forest_change)) %>%
  ggplot(aes(x = forest_change, y = entity, fill = entity)) +
  geom_col() +
  geom_vline(xintercept = 0, lty = 2) +
  scale_x_continuous(labels = scales::label_number(scale = 1/1e6, suffix = " MM")) +
  scale_fill_manual(values = single_highlight_pal) +
  expand_limits(x = c(-150e6, 150e6)) +
  labs(title = "By Continent",
       x = "Forest Area (hectares)",
       y = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(family = "Oswald", size = 16, hjust = 0.5),
        axis.title.x = element_text(size = 10))

p1
```

In the above figure, we see that South America has seen the biggest impact of deforestation with Africa being a relatively close second.  I decided to drill down further on the impact on countries making up South America.

```{r}
sa_countries <- c("Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", "Falkland Islands", "French Guyana",
                  "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela")

forest_change_sa <- forest_area_km %>%
  filter(entity %in% sa_countries) %>%
  filter(year == 1990 | year == 2020) %>%
  pivot_wider(names_from = year, names_prefix = "forest_area_", values_from = forest_area ) %>%
  mutate(forest_change = forest_area_2020 - forest_area_1990)

sa_highlight_pal <- c("Venezuela" = "#808080", "Colombia" = "#808080", "Argentina" = "#808080", 
                        "Bolivia" = "#808080", "Paraguay" = "#808080", "Brazil" = "#FFFF00")

p2 <- forest_change_sa %>%
    mutate(entity = fct_lump(entity, n = 6, w = abs(forest_change)),
           entity = fct_reorder(entity, forest_change)) %>%
  filter(entity != "Other") %>%
  ggplot(aes(x = forest_change, y = entity, fill = entity)) +
  geom_col() +
  geom_vline(xintercept = 0, lty = 2) +
  scale_x_continuous(labels = scales::label_number(scale = 1/1e6, suffix = " MM")) +
  scale_fill_manual(values = sa_highlight_pal) +
  expand_limits(x = c(-100e6, 100e6)) +
  labs(title = "By Country (SA)",
       x = "Forest Area (hectares)",
       y = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(family = "Oswald", size = 16, hjust = 0.5),
        axis.title.x = element_text(size = 10))

p2
```

Brazil was, by far, the biggest contributor to deforestation in South America compared to the other countries.  I suppose this result is not surprising given all of the media attention on loss of the Amazon rainforest.  The next analysis looked at different causes of deforestation in Brazil.

```{r}
brazil_cause <- brazil_loss %>%
  pivot_longer(cols = commercial_crops:small_scale_clearing, names_to = "cause", values_to = "forest_area") %>%
  group_by(cause) %>%
  summarize(forest_area = sum(forest_area), .groups = "drop") %>%
  mutate(total_forest = sum(forest_area)) %>%
  mutate(prop = forest_area / total_forest)

brazil_highlight_pal <- c("pasture" = "#FFFF00", "small_scale_clearing" = "#808080", "commercial_crops" = "#808080", 
                        "fire" = "#808080", "selective_logging" = "#808080", "tree_plantations_including_palm" = "#808080",
                        "natural_disturbances" = "#808080", "Other" = "#808080")

p3 <- brazil_cause %>%
  mutate(cause = fct_lump(cause, n = 7, w = prop),
         cause = fct_reorder(cause, prop, .desc = TRUE)) %>%
  ggplot(aes(x = cause, y = prop, fill = cause)) +
  geom_bar(stat = "identity", width = 0.5) + 
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_discrete(labels = function(x) str_wrap(
                              str_to_title(
                                str_replace_all(x, "_", " ")
                              ),10)
                            ) +
  scale_fill_manual(values = brazil_highlight_pal) +
  labs(title = "By Cause (Brazil)",
       x = NULL,
       y = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(family = "Oswald", size = 16, hjust = 0.5)
        )

p3
```

The main cause of deforestation in Brazil was found to be clearing land to create pasture for livestock grazing.  Finally, we combine all three plots into a single visualization using ```patchwork```.

```{r}
(p1 + p2) / p3 +
  plot_annotation(title = "Global Deforestation (1990-2020)",
                  caption = "Graphic: @datadavidz | Source: Our World in Data | #TidyTuesday",
                  theme = theme(
                    plot.title = element_text(family = "Oswald", size = 20),
                    plot.caption = element_text(size = 8))
                  )
```

### Summary

The main theme of this week's effort was to play around with highlighting on bar charts to divert the viewer's attention to certain data points.  Overall, I felt it was an effective technique for communicating the main sources of global deforestation.
