---
output: 
  html_document:
    code_download: true
    includes:
      after_body: footer.html
---
<br>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **TidyTuesday: HBCU Enrollments Dataset**
*Posted on February 2, 2021*

A quick analysis of the weekly [#TidyTuesday](http://github.com/rfordatascience/tidytuesday) dataset organized by the R4DS Online Learning Community.  My approach is to apply my data science skills to explore one question I have about the data and generate a visualization that addresses this question.  The main purpose for me is to practice and try out new things.  I am never completely satisfied with the end result but I do the best I can in a short period of time.

```{r, include=FALSE}

library(tidyverse)
library(tidytuesdayR)
library(waffle)

theme_set(theme_minimal())
```

**What I learned this week about R and the Tidyverse**

* My first use of the waffle package to make a waffle plot using geom_waffle.
* It is tricky to get the scales correct in a waffle bar plot.  Need to multiply back by a factor related to number of waffle rows.
* Used the truncated division operator (%/%) in function to assign years to decades.

**Brief explanation of the dataset**

The dataset contained number of enrollments at historically-black universities and colleges (HBCUs) from 1976-2015.  The main content of the dataset was the number of enrollments broken down by gender, public or private school and 2 or 4 year program.  My question was: "How did the enrollments change by HBCU type comparing 1990s vs. 2000s?".  Complete data was only available for these two decades.

```{r Load, include = FALSE}
#Save needed data into rds for blog post
hbcu_all <- readRDS(file = "./data/tt_210202.rds")
```

### Wrangle

Explore the data and process it into a nice format for plotting! Access each dataset by name by using a dollarsign after the `tt` object and then the name of the data set.

Initial conclusions from exploring the dataset:

* Years span from 1976-2015 but only 1990s and 2000s have complete yearly data.
* Four categories of HBCU: 2- and 4- year and Public or Private.

```{r Wrangle}
hbcu_enrollment <- hbcu_all %>%
  pivot_longer(`Total enrollment`:`2-year - Private`, names_to = "category", values_to = "enrollments")

hbcu_by_decade <- hbcu_enrollment  %>%
  filter(category %in% c("2-year - Private", "2-year - Public", "4-year - Private", "4-year - Public")) %>%
  filter(Year > 1989 & Year < 2010) %>%
  mutate(decade = paste0(10 * (Year %/% 10), "s")) %>%
  group_by(decade, category) %>%
  summarize(enrollments = sum(enrollments), .groups = "keep") %>%
  ungroup()
```
### Visualize

Initial visualization to look at enrollments across all data categories.
```{r Visualize}
hbcu_enrollment %>%
  mutate(name = fct_reorder(category, enrollments, last, .desc = TRUE)) %>%
  ggplot(aes(x = Year, y = enrollments, color = category)) +
  geom_line() +
  scale_y_continuous(labels = scales::comma_format())
```

Would like to create a waffle plot for each decade (1980s - 2010s) of Public, Private, 2-year, 4-year HBCUs.
```{r}
hbcu_by_decade %>%
  ggplot(aes(fill = category, values = enrollments %/% 10000)) +
  geom_waffle(color = "white", size = 0.25, n_rows = 10) +
  facet_wrap(~decade, ncol = 1, strip.position = "left") + 
  scale_x_continuous(labels = scales::unit_format(unit = "M", scale = 0.1)) +
  scale_y_discrete() +
  ggthemes::scale_fill_tableau(name=NULL) +
  labs(title = "HBCU Enrollments: 1990s vs. 2000s",
       subtitle = "Total Number of Enrollments (in Millions)",
       caption = "Visualization by @datadavidz | Data Courtesy of Data.World | #TidyTuesday") +
  ggthemes::theme_fivethirtyeight() +
  theme(panel.grid.major.x = element_blank(), axis.ticks.x = element_line())

```

### Summary

The number of HBCU enrollments increased from the 1990s to the 2000s largely driven by increase in the number of 2-year, public HBCU enrollments.  A function was required to adjust the labels so that the number of enrollments was listed accurately.  The scales package was also used to format the labels.  I used the "fivethirtyeight" theme but took the fill color from "tableau" both in the ggthemes package.
